---
title: "Human Genome Analysis Lab 10 : Data Maps, Interactive Graphs and Animations from the COVID-19 reporting data"
output:
  html_document:
    toc: true
    toc_depth : 4
    theme: united
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


```{r, warmings=FALSE, message=FALSE}
library(tidyverse)
library(maps)
library(mapdata)
library(lubridate)
library(viridis)
library(wesanderson)
```


### Preparing the times series data

The time series data is ripe for animation but first we need to get and format the files

```{r, warnings=FALSE, message=FALSE}
time_series_confirmed_long <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")  %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") 

# Let's get the times series data for deaths

time_series_deaths_long <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")  %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long),
               names_to = "Date", values_to = "Deaths")

time_series_recovered_long <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region") %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long),
               names_to = "Date", values_to = "Recovered")

# Create Keys 

time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)

time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)

time_series_recovered_long <- time_series_recovered_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Recovered)

# Join tables
time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key"))

time_series_long_joined <- full_join(time_series_long_joined,
              time_series_recovered_long, by = c("Key")) %>% 
    select(-Key)

# Reformat the data
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)

# Create Report table with counts
time_series_long_joined_counts <- time_series_long_joined %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long, Date),
               names_to = "Report_Type", values_to = "Counts")

```

### Creating the animations

Below are the packages I installed. There may be others that you need, in particular to rendering gifs. Several people needed to install the packages __gifski__ and __av__ Some of the examples may take several minutes to create the animation.

```{r, warnings=FALSE, message=FALSE}
library(ggplot2)
library(gganimate)
library(transformr)
theme_set(theme_bw())
```


```{r, warnings=FALSE, message=FALSE}
data_time <- time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths", "Recovered"), sum) %>% 
    filter (Country_Region %in% c("Italy","Spain", "France","China","United Kingdom","US", "Germany")) 

p <- ggplot(data_time, aes(x = Date,  y = Confirmed, color = Country_Region)) + 
      geom_point() +
      geom_line() +
      ggtitle("Confirmed Cases") 
p

```


```{r, warnings=FALSE, message=FALSE}
data_time <- time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths", "Recovered"), sum) %>% 
    filter (Country_Region %in% c("Italy","Spain", "France","China","United Kingdom","US", "Germany")) 

p <- ggplot(data_time, aes(x = Date,  y = Deaths, color = Country_Region)) + 
      geom_point() +
      geom_line() +
      ggtitle("Deaths") 
p

```


```{r}
time_series_long_joined_counts %>% 
  group_by(Country_Region, Report_Type, Date) %>% 
  summarise(Counts = sum(Counts)) %>% 
  filter (Country_Region == "US") %>% 
  filter (Report_Type %in% c("Confirmed", "Deaths")) %>% 
    ggplot(aes(x = Date,  y = Counts, fill = Report_Type, color = Report_Type)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Cases")
```

World Heat Map

```{r}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/04-13-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
# Fix mapping to map_data of US != USA  
  mutate(Country_Region = recode(Country_Region, US = "USA")) %>% 
  group_by(Country_Region) %>% 
  summarize(Confirmed = sum(Confirmed))
#  mutate(Country_Region = tolower(Country_Region))
# load the US map data
us <- map_data("state")

world <- map_data("world")
# We need to join the us map data with our daily report to make one data frame/tibble
country_join <- left_join(world, daily_report, by = c("region" = "Country_Region"))

# plot state map
ggplot(data = world, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add dasfdsta layer
  geom_polygon(data = country_join, aes(fill = Confirmed), color = "black") +
  scale_fill_gradientn(colours = 
                         wes_palette("Zissou1", 100, type = "continuous"),
                         trans = "log10") +
  labs(title = "COVID-19 Confirmed Cases in the world")
```



Pick a date

```{r}
picked_date <- time_series_long_joined %>% 
# Fix mapping to map_data of US != USA  
  mutate(Country_Region = recode(Country_Region, US = "USA")) %>% 
  filter(Date == "2020-04-01") %>% 
  group_by(Country_Region) %>% 
  summarize(Confirmed = sum(Confirmed))
#  mutate(Country_Region = tolower(Country_Region))
# load the US map data
us <- map_data("state")

world <- map_data("world")
# We need to join the us map data with our daily report to make one data frame/tibble
country_join <- left_join(world, picked_date, by = c("region" = "Country_Region"))

# plot state map
ggplot(data = world, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add dasfdsta layer
  geom_polygon(data = country_join, aes(fill = Confirmed), color = "black") +
  scale_fill_gradientn(colours = 
                         wes_palette("Zissou1", 100, type = "continuous"),
                         trans = "log10") +
  labs(title = "COVID-19 Confirmed Cases in the world")
```

