---
title: "Human Genome Analysis Lab 9 : Working with COVID-19 reporting data  "
output:
  html_document:
    toc: true
    toc_depth : 4
    theme: united
    highlight: tango
editor_options: 
  chunk_output_type: console
---

*I recognize, and fully understand, that this data maybe emotionally difficult to work. My intention is to make these lab relevant, allowing you to gather your own insights directly from new visualizations of the data. Please let me know if you would rather not work with the data.*

## Learning Objectives

* Understanding the sources of SARS-CoV-2 incidence reports
* Accessing data remotely
* Learn the difference between wide and long table formats
* Mapping data using latitude and longitude coordinates

## Visualizing COVID-19 cases, deaths and recoveries

The virus has been recently renamed based on phylogenetic analysis (more on this next week) severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2). The disease caused by the virus is coronavirus disease (COVID-19). In this lab we will work with reporting data on COVID-19 cases, deaths and recoveries. 

### Introduction to JHU case tracking data

Researchers (Ensheng Dong, Hongru Du, Lauren Gardner) at John Hopkins University developed an [interactive dashboard](https://www.arcgis.com/apps/opsdashboard/index.html#/bda7594740fd40299423467b48e9ecf6) to visual data and track reported cases of coronavirus disease 2019 (SARS-CoV-2) in real time. The underlying data is collated from the following sources and updated several times a day.

[World Health Organization (WHO)](https://www.who.int/) 
 | [DXY.cn. Pneumonia. 2020](http://3g.dxy.cn/newh5/view/pneumonia) 
 | [BNO News](https://bnonews.com/index.php/2020/02/the-latest-coronavirus-cases/)
 | [National Health Commission of the People’s Republic of China (NHC)](http://www.nhc.gov.cn/xcs/yqtb/list_gzbd.shtml) 
 | [China CDC (CCDC)](http://weekly.chinacdc.cn/news/TrackingtheEpidemic.htm)
 | [Hong Kong Department of Health](https://www.chp.gov.hk/en/features/102465.html)
 | [Macau Government](https://www.ssm.gov.mo/portal/)
 | [Taiwan CDC](https://sites.google.com/cdc.gov.tw/2019ncov/taiwan?authuser=0)
 | [US CDC](https://www.cdc.gov/coronavirus/2019-ncov/index.html)
 | [Government of Canada](https://www.canada.ca/en/public-health/services/diseases/coronavirus.html) 
 | [Australia Government Department of Health](https://www.health.gov.au/news/coronavirus-update-at-a-glance) 
 | [European Centre for Disease Prevention and Control (ECDC)](https://www.ecdc.europa.eu/en/geographical-distribution-2019-ncov-cases) 
 | [Ministry of Health Singapore (MOH)](https://www.moh.gov.sg/covid-19)
 | [Italy Ministry of Health](http://www.salute.gov.it/nuovocoronavirus)
 | [1Point3Arces](https://coronavirus.1point3acres.com/en)
 | [WorldoMeters](https://www.worldometers.info/coronavirus/)

* It is important to understand that this data is only as accurate as the reporting and many cases of the disease go unreported because of a lack of testing. This some countries may have have confirmed cases because of more comprehensive testing. Thus, the reporting data represent a minimum number of cases. See recent posts and an article by UMass statistian [Prof. Nick Reich](https://twitter.com/reichlab)

### JHU's Github repo

JHU researchers make data that goes into the [dashboard](https://www.arcgis.com/apps/opsdashboard/index.html#/bda7594740fd40299423467b48e9ecf6) available on [Github repo for Novel Coronavirus (COVID-19) Cases](https://github.com/CSSEGISandData/COVID-19). In this lab we will work with this data. 

You could start by creating a new Rproj and cloning the JHU repo as you did in Lab 6. However, we will not need all this data and it is easier to access the individual files directly as some are updated daily 

Let's take a look at the files and the structure of data in the files.

* csse_covid_19_data
  + csse_covid_19_daily_reports
    - 03-11-2020.csv
    
Open up the file to look at the structure 
    
The file contains the columns 

Province/State 	Country/Region 	Last Update 	Confirmed 	Deaths 	Recovered 	Latitude 	Longitude

It is important to note that for some countries there is only one row, while for others (e.g. China and US) there are multiple rows representing different provinces or states. Thus, we will need to sum these rows to get a total count for the US and China when we make graphs. From experience in making this tutorial I know the Column names with __/__ will cause errors in ggplot (). 

Next look at the time series data which is a complilation of the daily reports for each reporting type (Confirmed, Deaths, Recovered)

* csse_covid_19_data
  + csse_covid_19_time_series
    - time_series_covid19_confirmed_global.csv

Province/State 	Country/Region 	Lat 	Long 	1/22/20 	1/23/20 	1/24/20...
      
This data is what is called wide format, where values of the variable being observed (Confirmed cases) are spread out across columns (Here: columns for each day). Another way of describing this is that there is more than one measurement per row. This wide format works well for data entry and sometimes works well for analysis but can be cumbersome when using R. The __tidyr__ package allows us to quickly switch between wide format and what is called long format where there is only 1 column for the cases.

### Making a graphs from the daily reports

However, we are going to work with the data using __tidyverse__ functions and use a the library __lubridate__ for reformatting the dates

```{r, message=FALSE}
library(tidyverse)
library(lubridate)
```
                 
Let's start by plotting data from our last class on 3/11/2020 (the file looked at above). We can read data directly from the Github site. 

```{r, message=FALSE}
report_03_11_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-11-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State")
```

Check the table properties to make sure the data imported as we expected

```{r}
head(report_03_11_2020)
str(report_03_11_2020)
```

* Note that data from older reports is in a different format than the new format just release yesterday.

```{r, message=FALSE}
report_03_23_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-23-2020.csv"))
```

```{r}
head(report_03_23_2020)
str(report_03_23_2020)
```

*  Once change is the column titles, so that we no longer required to __rename__ . Note the addition of new columns to work with including county "Admin2" level information within a state. You will need to use summarize to get state or country counts.

Let's start with a plot of confirmed cases in US states as of March 11th.

```{r, fig.height=10}
report_03_11_2020 %>% 
  filter (Country_Region == "US") %>% 
  ggplot(aes(x = Confirmed,  y = reorder(Province_State, Confirmed))) + 
    geom_point() +
    ggtitle("Confirmed cases for each US State") +
    ylab("Country/Region") +
    xlab("Confirmed Cases")
```

If we try to graph by Country_Region there are entries that have the same character (e.g. China and US) so we need to summarise the data we are interested in graphing

```{r}
report_03_11_2020 %>% 
  group_by(Country_Region) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  arrange(desc(Deaths))
```

There are 116 countries which are a lot to put on a single graph, so let's just start with the countries with the most cases. In past labs we have only made bar graphs from histogram data, but values can be used where height of the bar will represent the value in a column of the data frame. This is done by using stat="identity" instead of the default, stat="bin".

```{r}
report_03_11_2020 %>% 
  group_by(Country_Region) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  arrange(desc(Deaths)) %>% 
  slice(1:20) %>% 
  ggplot(aes(x = Deaths,  y = reorder(Country_Region, Deaths))) + 
    geom_bar(stat = 'identity') +
    ggtitle("The 20 countries with the most reported COV19-related deaths") +
    ylab("Country/Region") +
    xlab("Deaths")
```

### Working with the time series data

#### Data Wrangling

There is some data wrangling required for working with the time series data. As of March 26, 2020 it still used the old column headers so we need to rename them. I expect this will change soon.

Start by loading the most recent times series data for confirmed cases

```{r, message=FALSE}
time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```

Check the table properties to make sure the data imported as we expected

```{r}
head(time_series_confirmed)
```

As noted above this data is in wide format. To convert to long format

```{r}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") 
```

Let's look at the format of the data frame (tibble) now

```{r}
head(time_series_confirmed_long)
```

It would be convenient to have the confirmed cases and deaths in the same table. We can create another table with the deaths and then join the two tables. 

Let's get the times series data for deaths

```{r, message=FALSE}
time_series_deaths <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```

and convert it to long format

```{r}
time_series_deaths_long <- time_series_deaths %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long),
               names_to = "Date", values_to = "Deaths")
head(time_series_deaths_long)
```

To join 2 tables we need a common column in which each row is a unique name. This is often called the Key. None of the columns in this data set meet that criteria, but we can create a column to serve as our key by mergering the names for the Province_State, Country_Region and Date columns using __unite__

```{r}
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)
```

For the second table we can do the same and eliminate the columns that are redundant so that we just have the Key and Deaths columns

```{r}
time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)
```

Now to join the tables. A __full_join__ which means that the keys that are common to the tables will be joined and there will be counts for both Confirmed and Deaths. Where there are not matching values, returns NA for the one missing. __select__ can be used to remove the key after the join since we don't have further need for it.

```{r}
time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key")) %>% 
    select(-Key)
head(time_series_long_joined)
```

Check to make sure neither the Cofirmed or Death counts have NA as a value (both tables have the same number of rows, but we should check to make sure)

```{r}
which(is.na(time_series_long_joined$Confirmed))
which(is.na(time_series_long_joined$Deaths))
```

There are no values so we are in good shape to proceed to using this table for graphing, but first reformat Date

```{r}
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)
```

#### Make Graphs from the time series data

To make a times series graph of the confirmed cases we need to summarize the Country date to count up the individual state data for the US. 

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise(Confirmed = sum(Confirmed)) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = Confirmed)) + 
    geom_point() +
    geom_line() +
    ggtitle("US Confirmed COVID-19 Cases")
```

Let's add in a few other countries

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise(Confirmed = sum(Confirmed)) %>% 
  filter (Country_Region %in% c("China","Japan", "Korea, South",
                                "Italy","Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Confirmed)) + 
    geom_point() +
    geom_line() +
    ggtitle("Confirmed COVID-19 Cases") +
    facet_wrap(~Country_Region, ncol=2, scales="free_y")
```

Now several countries on the same graph

```{r}
time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise(Confirmed= sum(Confirmed)) %>% 
    filter (Country_Region %in% c("China","France","Italy", 
                                "Korea, South", "Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Confirmed, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths")
```

### Visualizing data on maps

Here are two examples provided by Prof. Chris Sunderland and Anisa Dhana

#### Adaptation of a post from Anisa Dhana on [R-Bloggers](https://www.r-bloggers.com/map-visualization-of-covid-19-across-the-world-with-r/) 

* This code uses data is in wide format from the above times series.

Load the additional libraries (and install if necessary)

```{r, message=FALSE}
library(maps)
library(viridis)
```

Get the world map
```{r}
world <- map_data("world")
```

Cutoffs based on the number of cases
```{r}
mybreaks <- c(1, 20, 100, 1000, 50000)
```

Plot of data from March 11

```{r, warnings=FALSE}
ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point(data=time_series_confirmed, aes(x=Long, y=Lat, size=`2/25/20`, color=`2/25/20`),stroke=F, alpha=0.7) +
  scale_size_continuous(name="Cases", trans="log", range=c(1,7),breaks=mybreaks, labels = c("1-19", "20-99", "100-999", "1,000-49,999", "50,000+")) +
  # scale_alpha_continuous(name="Cases", trans="log", range=c(0.1, 0.9),breaks=mybreaks) +
  scale_color_viridis_c(option="inferno",name="Cases", trans="log",breaks=mybreaks, labels = c("1-19", "20-99", "100-999", "1,000-49,999", "50,000+")) +
  theme_void() + 
  guides( colour = guide_legend()) +
  labs(caption = "") +
  theme(
    legend.position = "bottom",
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "#ffffff", color = NA), 
    panel.background = element_rect(fill = "#ffffff", color = NA), 
    legend.background = element_rect(fill = "#ffffff", color = NA)
  )
```

#### Adaptation of code from Prof. Chris Sunderland

Get data, pivot to longer format and summarize
```{r, messages=FALSE}
time_series_confirmed_long2 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
    rename(Province_State = "Province/State", Country_Region = "Country/Region") %>%
    pivot_longer(-c(Province_State, Country_Region, Lat, Long),
    names_to = "Date", values_to = "cumulative_cases") %>%
    mutate(Date = mdy(Date) - days(1),
        Place = paste(Lat,Long,sep="_")) %>%
    group_by(Place,Date) %>%
        summarise(cumulative_cases = ifelse(sum(cumulative_cases)>0,
        sum(cumulative_cases),NA_real_),
        Lat = mean(Lat),
        Long = mean(Long)) %>%
    mutate(Pandemic_day = as.numeric(Date - min(Date)))
```

View the table
```{r}
head(time_series_confirmed_long2)
```

Plot the data
```{r, fig_height = 8, warnings=FALSE}
static <- ggplot(subset(time_series_confirmed_long2,Date %in% seq(min(Date),max(Date),14)),
            aes(x = Long, y = Lat, size = cumulative_cases/1000)) +
            borders("world", colour = NA, fill = "grey90") +
            theme_bw() +
            geom_point(shape = 21, color='purple', fill='purple', alpha = 0.5) +
            labs(title = 'COVID-19 spread',x = '', y = '',
                 size="Cases (x1000))") +
            theme(legend.position = "right") +
            coord_fixed(ratio=1)+
            facet_wrap(.~Date,ncol=1)
static
```

# Exercises

Create at least 2 new graphs (different from the ones above) each from the categories of daily reports, time series and geographical maps that best convey aspects of the pandemic relevant to you.

Submit this exercise as a link from your Github page (Submit your Github website URL in Moodle).

