---
title: "Human Genome Analysis Lab : Alternative to Genetic Testing Report and Presentation - Human Population Data"
output:
  html_document:
    toc: true
    toc_depth : 4
    theme: united
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

*I recognize, and fully understand, that this data maybe emotionally difficult to work. My intention is to make these lab relevant, allowing you to gather your own insights directly from new visualizations of the data. Please let me know if you would rather not work with the data.*

## Learning Objectives

* The goal of this lab is to integrate US population data with the COVID-19 reporting data in a Shiny App

## US Census Shiny App

In the RStudio Shiny tutorials [Lesson 5](https://shiny.rstudio.com/tutorial/written-tutorial/lesson5/)
.  Go through this tutorial to ensure you can create a Shiny App using the US census data package. 


## Integrating the US Census data with the COVID-19 reporting data

In [Lesson 5](https://shiny.rstudio.com/tutorial/written-tutorial/lesson5/) there is a step in which you download the census data (counties.rds) to your computer.

```{r}
counties <- readRDS("ShinyApps/data/counties.rds") # Note I used a different directory
head(counties)
```

If you go to [JHU times series reporting data](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series) you can also see that there is country level information on confirmed cases and deaths 

```{r, warnings=FALSE, message=FALSE}
library(tidyverse)
US_time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"))
head(US_time_series_confirmed, 10)
```

There is some data wrangling to be done to join the US Census data with the COVID-19 confirmed cases and deaths involving differences in capitalization, merging (or separating columns).  However, I think you are ready for the rodeo based on your accomplishments this semester! Remember there are examples using the covid data at the country level in Lab 10.

## Exercises

1. Create a Shiny App that allows the user to select (1) the date, (2) confirmed cases or deaths and (3) cases per county, cases divided by total population per county. Plot the results on the US county level map.

The goal of these exercises are to further develop your data wrangling and graphing skills using real world data. In previous labs we put most of the data wrangling steps upstream of the Shiny App ui and server code. This is prefered because these steps are only do once, not each time a new value or field is selected by the user in the ui or server sections.  

To date the approach for making Shiny Apps to create a single data frame/tibble with the needed data. Using the county map data in combination with census and the US time series covid-19 data generates large tables that can tax your computers RAM memory and cause R to crash. A way to reduce your data frame sizes is to include the user selection before you combine the date. For example in Exercise #1, filtering the time series data by date before joining the table decreases the table length nearly 100x.

For Exercise 1 you could start by

1. Load your libraries
2. Join county level census data and map data (see example in Lab 10 on working with the map data)
3. Join the US time series data for the Confirmed Cases and Deaths 
4. In the ui section get the date from the user
5. In the server section filter the Confirmed_Deaths data frame based on the date
6. In the server section join the Confirmed_Deaths and census_map data frames.
7. Use the above data frame in the graph

Here is an [example](https://jeffreyblanchard.shinyapps.io/Including_population_data/) that I made. Yours may look different.

Submit the url to your Shiny app in Moodle.



